---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# rb3

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Codecov test coverage](https://codecov.io/gh/wilsonfreitas/rb3/branch/main/graph/badge.svg)](https://app.codecov.io/gh/wilsonfreitas/rb3?branch=main)
[![R build (rcmdcheck)](https://github.com/wilsonfreitas/rb3/workflows/R-CMD-check/badge.svg)](https://github.com/wilsonfreitas/rb3/actions)

<!-- badges: end -->

Download and read a bunch o data from [B3](https://www.b3.com.br) making it easy
to store, process and consume it in a structure way.

# Documentation

The documentation is available [here](https://wilsonfreitas.github.io/rb3/),
the reference and articles with real applications can be found.

## Installation

```r
devtools::install_github("wilsonfreitas/rb3")
```

## Examples

### Yield curve

Download and use historical yield curve data with `yc_get`.

```{r setup}
library(rb3)
library(ggplot2)
library(stringr)

df_yc <- yc_get(
    first_date = Sys.Date() - 255 * 5,
    last_date = Sys.Date(),
    by = 255
)

p <- ggplot(
    df_yc,
    aes(
        x = forward_date,
        y = r_252,
        group = refdate,
        color = factor(refdate)
    )
) +
    geom_line() +
    labs(
        title = "Yield Curves for Brazil",
        subtitle = "Built using interest rates future contracts",
        caption = str_glue("Data imported using rb3 at {Sys.Date()}"),
        x = "Forward Date",
        y = "Annual Interest Rate",
        color = "Reference Date"
    ) +
    theme_light() +
    scale_y_continuous(labels = scales::percent)

print(p)
```

### Futures prices

Get settlement future prices with `futures_get`.

```{r message=FALSE}
library(rb3)
library(dplyr)

df <- futures_get(
    first_date = "2022-04-01",
    last_date = "2022-04-29",
    by = 5
)

df |> filter(commodity == "DI1")
```

### Equity data

Equity closing data (without price adjustments) is available thru `cotahist_get`.

```{r}
library(rb3)
library(bizdays)

date <- preceding(Sys.Date(), "Brazil/ANBIMA") # last business day
ch <- cotahist_get(date, "daily")
cotahist_equity_get(ch)
```

Funds data

```{r}
cotahist_funds_get(ch)
```

BDRs data

```{r}
cotahist_bdrs_get(ch)
```

Equity options

```{r}
cotahist_equity_options_get(ch)
```
