---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rb3

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Codecov test coverage](https://codecov.io/gh/wilsonfreitas/rb3/branch/main/graph/badge.svg)](https://app.codecov.io/gh/wilsonfreitas/rb3?branch=main)
[![R build (rcmdcheck)](https://github.com/wilsonfreitas/rb3/workflows/R-CMD-check/badge.svg)](https://github.com/wilsonfreitas/rb3/actions)

[![CRAN status](https://www.r-pkg.org/badges/version/rb3)](https://CRAN.R-project.org/package=rb3)

[![](https://cranlogs.r-pkg.org/badges/rb3)](https://cran.r-project.org/package=rb3)

<!-- badges: end -->

[B3](https://www.b3.com.br)  is the main financial exchange in Brazil, offering 
support and access to trading systems for equity and fixed income markets. 
In its website you can find a vast number of datasets regarding prices and transactions 
for contracts available for trading at these markets. 

Package **rb3** facilitates downloading and reading these datasets from 
[B3](https://www.b3.com.br), making it easy to consume it in R in a structured way.


# Documentation

The documentation is available in its [pkgdown page](https://wilsonfreitas.github.io/rb3/),
where articles (vignettes) with real applications can be found.

## Installation

```r
install.packages("rb3")
```

```r
# github (Development branch)
if (!require(devtools)) install.packages("devtools")
devtools::install_github("wilsonfreitas/rb3")
```

## Examples

### Yield curve

Download and use historical yield curve data with `yc_get`.

```{r setup}
library(rb3)
library(ggplot2)
library(stringr)

df_yc <- yc_mget(
  first_date = Sys.Date() - 255 * 5,
  last_date = Sys.Date(),
  by = 255
)

p <- ggplot(
  df_yc,
  aes(
    x = forward_date,
    y = r_252,
    group = refdate,
    color = factor(refdate)
  )
) +
  geom_line() +
  labs(
    title = "Yield Curves for Brazil",
    subtitle = "Built using interest rates future contracts",
    caption = str_glue("Data imported using rb3 at {Sys.Date()}"),
    x = "Forward Date",
    y = "Annual Interest Rate",
    color = "Reference Date"
  ) +
  theme_light() +
  scale_y_continuous(labels = scales::percent)

print(p)
```

### Futures prices

Get settlement future prices with `futures_get`.

```{r message=FALSE}
library(rb3)
library(dplyr)

df <- futures_mget(
  first_date = "2022-04-01",
  last_date = "2022-04-29",
  by = 5
)

glimpse(
  df |>
    filter(commodity == "DI1")
)
```

### Equity data

Equity closing data (without **ANY** price adjustments) is available thru `cotahist_get`.

```{r}
library(rb3)
library(bizdays)

# fix for ssl error (only in linux)
if (Sys.info()["sysname"] == "Linux") {
  httr::set_config(
    httr::config(ssl_verifypeer = FALSE)
  )
}

date <- preceding(Sys.Date() - 1, "Brazil/ANBIMA") # last business day
ch <- cotahist_get(date, "daily")

glimpse(
  cotahist_equity_get(ch)
)
```

Funds data

```{r}
glimpse(
  cotahist_funds_get(ch)
)
```

BDRs data

```{r}
glimpse(
  cotahist_bdrs_get(ch)
)
```

Equity options

```{r}
glimpse(
  cotahist_equity_options_get(ch)
)
```
