---
title: "Fetching Historical Equity Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fetching Historical Equity Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

```{r load-data, echo=FALSE}
# This loads pre-saved data to avoid network calls during CRAN checks
# See vignette creation script in data-raw/vignettes/save_equity_data.R
load("equity_data.RData")
```

## Introduction

The B3 (Brasil, Bolsa, Balcão) provides historical stock market data through its COTAHIST files.
These files contain valuable information about stocks, ETFs, BDRs, UNITs, REITs (FIIs), FIAGROs,
and FIDCs trading on the Brazilian exchange.
The `rb3` package simplifies access to this data through the templates `b3-cotahist-yearly` and
`b3-cotahist-daily`, which handle the downloading and processing of these files.

The COTAHIST files represent one of the most comprehensive sources of historical market data in
Brazil, with annual files available as far back as 1986.
However, due to Brazil's monetary stabilization process in 1994 (Plano Real),
it is generally recommended to use data from 1995 onwards.

It's important to note that the prices in these files are not adjusted for corporate actions,
which means that for long-term historical analysis of individual stocks,
additional adjustments may be necessary.

In this vignette, we'll explore how to fetch, filter, and analyze historical equity data using
the `rb3` package.

```{r setup, message=FALSE, warning=FALSE}
library(rb3)
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Fetching historical data

To access historical equity data, we need to download the COTAHIST files and create the
corresponding datasets.
The `rb3` package provides functions to seamlessly handle this process.

### Downloading yearly data

For comprehensive historical analysis, you can download yearly COTAHIST files.
This is ideal for long-term studies covering multiple years.

```{r download-yearly, eval=FALSE}
# Download data for multiple years (example: 2018 to 2024)
fetch_marketdata("b3-cotahist-yearly", year = 2018:2024)
```

### Downloading daily data

For more recent analysis or when you need specific dates, you can fetch daily data:

```{r download-daily, eval=FALSE}
# Download data for specific dates
fetch_marketdata("b3-cotahist-daily", refdate = as.Date("2024-01-02"))
```

### Accessing the datasets

After downloading the data, you can access it using the `cotahist_get()` function:

```{r access-data, eval=FALSE}
# Access yearly dataset
ch <- cotahist_get("yearly")

# Access daily dataset
ch_daily <- cotahist_get("daily")
```

The `cotahist_get()` function returns a lazy query that doesn't immediately load the data into
memory.
This approach allows efficient manipulation of large datasets.
To materialize the data as an R tibble, use the `collect()` function.

## Filtering by asset type

The COTAHIST files contain data for various types of assets.
The `rb3` package provides specialized filter functions to extract specific asset classes:

### Equities

For stocks trading on B3:

```{r equity-filter, eval=FALSE}
# Filter for stocks from 2024
eq <- ch |>
  filter(year(refdate) == 2024) |>
  cotahist_filter_equity()
```

### ETFs (Exchange Traded Funds)

For ETFs listed on B3:

```{r etf-filter, eval=FALSE}
# Filter for ETFs from 2024
etfs <- ch |>
  filter(year(refdate) == 2024) |>
  cotahist_filter_etf()
```

### REITs (FIIs - Fundos Imobiliários)

For real estate investment funds:

```{r fii-filter, eval=FALSE}
# Filter for REITs from 2024
fii <- ch |>
  filter(year(refdate) == 2024) |>
  cotahist_filter_fii()
```

### BDRs (Brazilian Depositary Receipts)

For Brazilian depositary receipts of international companies:

```{r bdr-filter, eval=FALSE}
# Filter for BDRs
bdrs <- cotahist_get() |>
  cotahist_filter_bdr()
```

### Other asset classes

Additional filter functions are available for other asset types:

- `cotahist_filter_unit()`: For UNITs (sets of securities traded as a single unit)
- `cotahist_filter_fidc()`: For FIDCs (Credit Rights Investment Funds)
- `cotahist_filter_fiagro()`: For FIAGROs (Agribusiness Investment Funds)
- `cotahist_filter_index()`: For market indices
- `cotahist_filter_equity_options()`: For stock options
- `cotahist_filter_index_options()`: For index options
- `cotahist_filter_fund_options()`: For fund options

## Analyzing equity data

Let's explore some common analyses you might perform with this historical data.

### Top traded equities by volume

This example identifies the most actively traded stocks and visualizes their monthly trading
volumes:

```{r equity-volume-fake, eval=FALSE}
# Get top 10 stocks by volume
symbols_eq <- eq |>
  group_by(symbol) |>
  summarise(volume = sum(volume)) |>
  arrange(desc(volume)) |>
  head(10) |>
  pull(symbol, as_vector = TRUE)

# Calculate monthly volumes for top stocks
eq_monthly <- eq |>
  filter(symbol %in% symbols_eq) |>
  mutate(month = floor_date(refdate, "month")) |>
  group_by(month, symbol) |>
  summarise(volume = sum(volume)) |>
  collect()
```

```{r equity-volume-plot, echo=FALSE, fig.width=7, fig.height=5.5}
# Using pre-loaded data to create the plot without fetching data
ggplot(eq_monthly, aes(x = month, y = volume, group = symbol, colour = symbol)) +
  geom_line() +
  scale_y_continuous(labels = scales::label_number()) +
  labs(
    title = "Monthly Trading Volume for Top 10 Brazilian Stocks",
    x = "Month",
    y = "Volume (BRL)",
    caption = "Data source: B3 via rb3 package"
  )
```

### ETF market share analysis

The market share of different ETFs provides insights into investor preferences:

```{r etf-analysis-fake, eval=FALSE}
# Calculate total ETF volume
total_volume <- etfs |>
  summarise(volume = sum(volume)) |>
  pull(volume, as_vector = TRUE)

# Get volume share for top ETFs
etf_shares <- etfs |>
  group_by(symbol) |>
  summarise(volume = sum(volume)) |>
  collect() |>
  mutate(volume_ratio = volume / total_volume) |>
  slice_max(volume_ratio, n = 10) |>
  mutate(volume_ratio_acc = cumsum(volume_ratio))
```

```{r etf-share-plot, echo=FALSE, fig.width=7, fig.height=5.5}
# Using pre-loaded data to create the plot
fmt <- scales::label_percent(accuracy = 0.1)

ggplot(etf_shares, aes(
  x = reorder(symbol, -volume_ratio), y = volume_ratio,
  label = fmt(volume_ratio)
)) +
  geom_bar(stat = "identity", fill = "royalblue") +
  geom_text(nudge_y = 0.01) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    x = NULL, y = NULL,
    title = "Volume Share of Top 10 ETFs",
    subtitle = "Percentage of Total ETF Trading Volume",
    caption = "Data source: B3 via rb3 package"
  )
```

### Analyzing distribution IDs for price continuity

The `distribution_id` field helps identify changes in a stock's pricing series,
which might be due to corporate actions:

```{r distribution-analysis-fake, eval=FALSE}
# Select a specific stock
symbol_ <- "ITUB4"

# Extract data for the selected stock
stock_data <- equities |>
  filter(symbol == symbol_) |>
  arrange(refdate) |>
  collect()
```

```{r distribution-id-plot, echo=FALSE, fig.width=7, fig.height=5}
# Using pre-loaded data to create the plot
ggplot(stock_data, aes(x = refdate, y = distribution_id)) +
  geom_line() +
  labs(
    title = "Distribution ID Changes for ITUB4",
    x = "Date",
    y = "Distribution ID",
    caption = "Data source: B3 via rb3 package"
  )
```

```{r price-by-distribution-plot, echo=FALSE, fig.width=7, fig.height=5}
# Using pre-loaded data to create the plot
ggplot(stock_data, aes(x = refdate, y = close, colour = factor(distribution_id))) +
  geom_line() +
  labs(
    title = "Price History for ITUB4 by Distribution ID",
    x = "Date",
    y = "Closing Price (BRL)",
    color = "Distribution ID",
    caption = "Data source: B3 via rb3 package"
  )
```

## Working with BDRs

Brazilian Depositary Receipts (BDRs) represent shares of companies listed on foreign exchanges:

```{r bdr-analysis-fake, eval=FALSE}
# Find the most common BDRs in the dataset
top_bdrs <- bdrs |>
  group_by(symbol) |>
  count(sort = TRUE) |>
  head(15) |>
  collect()

# Analyze trading volume for a specific BDR
bdr_data <- bdrs |>
  filter(symbol == "AAPL34") |>
  arrange(refdate) |>
  collect()
```

```{r bdr-volume-plot, echo=FALSE, fig.width=7, fig.height=5}
# Using pre-loaded data to create the plot
ggplot(bdr_data, aes(x = refdate, y = trade_quantity, colour = factor(distribution_id))) +
  geom_line() +
  labs(
    title = "Trading Volume for AAPL34 (Apple BDR)",
    x = "Date",
    y = "Trade Quantity",
    color = "Distribution ID",
    caption = "Data source: B3 via rb3 package"
  )
```

## Conclusion

In this vignette, we've explored how to fetch and analyze historical equity data from B3 using
the `rb3` package.
The package provides efficient tools for downloading COTAHIST files and extracting specific asset
classes through specialized filter functions.

The key steps in working with historical equity data include:

1.  **Downloading data** using `fetch_marketdata()` with either the `b3-cotahist-yearly` or
    `b3-cotahist-daily` templates
2.  **Accessing datasets** with `cotahist_get()`
3.  **Filtering by asset type** using specialized functions like `cotahist_filter_equity()`,
    `cotahist_filter_etf()`, etc.
4.  **Analyzing the data** using standard R data manipulation and visualization tools

This workflow enables a wide range of analyses, from studying trading volumes and market shares to
tracking price movements and identifying corporate actions through distribution IDs.

For more advanced analyses, you might consider:
- Adjusting prices for corporate actions
- Building portfolios and calculating returns
- Analyzing liquidity patterns
- Studying sector performance
- Constructing custom indices

The combination of B3's comprehensive historical data and the `rb3` package's user-friendly
interface provides a powerful toolkit for researching and analyzing the Brazilian equity market.