---
title: "Fetching Market Data with rb3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fetching Market Data with rb3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `rb3` package provides tools for downloading, processing, and analyzing market data from B3 (the Brazilian stock exchange). One of the key functions in the package is `fetch_marketdata()`, which simplifies the process of downloading and storing market data in a structured format.

This vignette demonstrates how to use the `fetch_marketdata()` function to retrieve various types of market data from B3.

```{r setup}
library(rb3)
library(dplyr)
library(ggplot2)
library(bizdays)
```

## Understanding `fetch_marketdata()`

The `fetch_marketdata()` function performs two main steps:

1. Downloads market data files by creating all combinations of the provided parameters and calling `download_marketdata()` for each combination
2. Processes the downloaded files by reading them into a database using `read_marketdata()`

The function has the following signature:

```r
fetch_marketdata(template, ...)
```

Where:
- `template`: A character string specifying the market data template to use
- `...`: Named arguments that will be expanded into a grid of all combinations to fetch data for

## Available Templates

You can view all available templates using the `list_templates()` function:

```{r}
list_templates()
```

The most commonly used templates include:

- `b3-cotahist-daily`: Daily stock market historical quotes
- `b3-cotahist-yearly`: Yearly stock market historical quotes
- `b3-futures-settlement-prices`: Settlement prices for futures contracts
- `b3-reference-rates`: Reference interest rates

## Examples

### Fetching Daily Historical Data

To download daily historical data for a specific date range:

```{r}
# Download daily historical data for a date range
fetch_marketdata(
  "b3-cotahist-daily", 
  refdate = bizseq("2023-01-01", "2023-01-31", "Brazil/B3")
)

# Now the data is available in the database
stocks <- cotahist_get(
  refdate = as.Date("2023-01-15"),
  type_market = "VISTA"
) |> 
  collect()

# Display the top 10 stocks by trading volume
stocks |>
  filter(type_stock == "ON") |>
  arrange(desc(volume)) |>
  head(10) |>
  select(symbol, name, price_close, volume)
```

### Fetching Yearly Historical Data

For accessing historical data for multiple years:

```{r}
# Download yearly historical data for 2020-2023
fetch_marketdata("b3-cotahist-yearly", year = 2020:2023)

# Analyze market data from 2020
yearly_data <- cotahist_get(
  year = 2020,
  type_market = "VISTA"
) |> 
  collect()

# Calculate monthly returns for a specific stock
monthly_returns <- yearly_data |>
  filter(symbol == "PETR4") |>
  mutate(month = format(refdate, "%Y-%m")) |>
  group_by(month) |>
  summarize(
    open = first(price_open),
    close = last(price_close),
    return = (close / open) - 1
  )
```

### Fetching Futures Settlement Prices

To download futures settlement prices for a specific time period:

```{r}
# Download futures settlement prices for mondays in January 2023
mondays <- getdate("monday", as.Date("2023-01-01") + 0:30, "Brazil/B3")
fetch_marketdata("b3-futures-settlement-prices", refdate = mondays)

# Access futures data
futures_data <- futures_get() |> 
  filter(commodity == "DI1") |>
  collect()

# Plot futures prices
futures_data |>
  filter(symbol %in% c("DI1F24", "DI1F25")) |>
  ggplot(aes(x = refdate, y = price, color = symbol)) +
  geom_line() +
  labs(
    title = "DI1 Futures Prices",
    x = "Date",
    y = "Price",
    color = "Contract"
  )
```

### Fetching Reference Rates

To download reference interest rates for various curves:

```{r}
# Download reference rates for multiple curves and dates
fetch_marketdata(
  "b3-reference-rates", 
  refdate = bizseq("2023-01-01", "2023-03-31", "Brazil/B3"),
  curve_name = c("DIC", "DOC", "PRE")
)

# Access PRE curve data (nominal interest rates)
pre_curve <- yc_brl_get() |>
  filter(refdate >= as.Date("2023-01-01"), refdate <= as.Date("2023-03-31")) |>
  collect()

# Plot yield curves for different dates
pre_curve |>
  filter(forward_date < as.Date("2030-01-01")) |>
  mutate(ref_month = format(refdate, "%Y-%m")) |>
  filter(ref_month %in% c("2023-01", "2023-02", "2023-03")) |>
  ggplot(aes(x = forward_date, y = r_252, color = factor(refdate))) +
  geom_line() +
  labs(
    title = "PRE Yield Curves",
    x = "Forward Date",
    y = "Annual Interest Rate",
    color = "Reference Date"
  ) +
  scale_y_continuous(labels = scales::percent)
```

## Combining Multiple Parameter Sets

One of the powerful features of `fetch_marketdata()` is its ability to handle multiple parameter combinations. The function automatically creates a grid of all possible combinations and processes each one:

```{r}
# Download data for multiple dates and curves simultaneously
fetch_marketdata(
  "b3-reference-rates",
  refdate = c(as.Date("2023-01-15"), as.Date("2023-06-15")),
  curve_name = c("DIC", "PRE")
)
```

This will download 4 different datasets (2 dates × 2 curves).

## Performance Considerations

When fetching large amounts of data, keep in mind:

1. The function shows progress indicators for both downloading and processing steps
2. Data is cached locally, so subsequent queries for the same data are faster
3. The database storage uses efficient Parquet format for optimal query performance
4. For very large datasets, consider fetching data in smaller batches

## Conclusion

The `fetch_marketdata()` function provides a convenient way to download, process, and store market data from B3. By combining it with the other query functions in the `rb3` package, you can easily access and analyze Brazilian market data for research, trading strategies, or risk management.

Remember that all downloaded data is stored in a local database managed by the `rb3` package, making it readily available for queries without needing to download it again.