---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `rb3` package provides tools for downloading, processing, and analyzing market data from B3
(the Brazilian stock exchange). This vignette will guide you through the basics of using the
package to download various types of market data and perform common analyses.

```{r setup, eval=TRUE, message=FALSE}
library(rb3)
library(dplyr)
library(bizdays)
```

## Downloading Market Data

The main function for fetching market data is `fetch_marketdata()`.
This function downloads data based on a template and parameter combinations, then processes
the data into a structured database format.

### Templates

Templates are predefined configurations that specify the type of data to download and how to
process it.
Each template corresponds to a specific dataset or file type available from B3.
For example:

- `"b3-cotahist-yearly"`: Downloads and reads COTAHIST file that are available by year.
- `"b3-futures-settlement-prices"`: Downloads and reads settlement prices web page.
- `"b3-reference-rates"`: Downloads and reades the web page of reference interest rates.
- `"b3-bvbg-086"`: Downloads and reads the BVBG-086 file with trading instruments information.

```{r list-templates, eval=TRUE}
# List available templates
list_templates()
```

Additional information about templates can be obtained by calling the `template_retrieve()`
function:

```{r template-retrieve}
# Get a specific template
template_retrieve("b3-cotahist-yearly")
#> Template: b3-cotahist-yearly
#> Description: Cotações Históricas do Pregão de Ações - Arquivo Anual
#> Required arguments:
#> • year: Ano de referência
#> Fields:
#> • regtype (numeric): Tipo de registro
#> • refdate (Date): Data do pregão
#> • bdi_code (numeric): Código BDI
#> • symbol (character): Código de negociação do papel
#> • instrument_market (numeric): Tipo de mercado
#> • corporation_name (character): Nome resumido da empresa emissora do papel
#> • specification_code (character): Especificação do papel
#> • days_to_settlement (numeric): Prazo em dias do mercado a termo
#> • trading_currency (character): Moeda de referência
#> • open (numeric): Preço de abertura do papel
#> • high (numeric): Preço máximo do papel
#> • low (numeric): Preço mínimo do papel
#> • average (numeric): Preço médio do papel
#> • close (numeric): Preço último negócio efetuado com o papel
#> • best_bid (numeric): Preço da melhor oferta de compra do papel
#> • best_ask (numeric): Preço da melhor oferta de venda do papel
#> • trade_quantity (numeric): Número de negócios efetuados com o papel
#> • traded_contracts (numeric): Quantidade total de títulos negociados neste
#> papel
#> • volume (numeric): Volume total de títulos negociados neste papel
#> • strike_price (numeric): Preço de exercício para o mercado de opções ou valor
#> do contrato para o mercado de termo secundário
#> • strike_price_adjustment_indicator (character): Indicador de correção de
#> preços de exercícios ou valores de contrato para os mercados de opções, termo
#> secundário ou futuro
#> • maturity_date (Date): Data do vencimento para os mercados de opções, termo
#> secundário ou futuro
#> • allocation_lot_size (numeric): Fator de cotação do papel
#> • strike_price_in_points (numeric): Preço de exercício em pontos para opções
#> referenciadas em dólar ou valor de contrato em pontos para termo secundário
#> • isin (character): Código do papel no sistema ISIN
#> • distribution_id (numeric): Número de distribuição do papel
```

Once you know the template you want to use, you can download the data by calling
`fetch_marketdata()`.
The function takes the template name and additional parameters as arguments.

### Fetching market data

The `fetch_marketdata()` function downloads and processes market data based on the specified
template and parameters.
The data is stored in a local database, which can be queried using specialized functions.
The code below shows an example on how to download and process data using the `fetch_marketdata()`.

```{r fetch-reference-rates, eval=FALSE}
# Download daily historical data for a specific date range
fetch_marketdata("b3-reference-rates",
  refdate = bizseq("2024-01-01", "2024-01-31", "Brazil/B3"),
  curve_name = c("PRE", "DIC")
)
#> ✔ Downloading data [53s]
#> ℹ 44 files downloaded
#> ✔ Reading data into DB [6s]
```

The total time taken to download and process the data is shown in the console output,
and also the number of downloaded files is shown.
This code downloads 44 files containing reference rates for the PRE and DIC curves for January 2024.
The files are read and stored as parquet files forming a local database inside the `rb3.cachedir`
folder.

### `rb3.cachedir` folder

The `rb3.cachedir` folder is where the downloaded data is stored.
It is set as an option in R, and you can check its current value using:

```{r view-rb3-cachedir}
getOption("rb3.cachedir")
#> [1] "/home/wilson/dev/rb3/rb3-cache"
```

You can change the location of the `rb3.cachedir` folder by setting the option
`rb3.cachedir` to a different path.
```{r set-rb3-cachedir, eval=FALSE}
# Set the rb3.cachedir folder to a different path
options(rb3.cachedir = "/path/to/your/custom/folder")
```

> #### Note
> It is strongly recommended to set the `rb3.cachedir` folder in the .Rprofile file.

Inside this folder it has the 3 folders:

- raw: for raw downloaded files
- meta: with the metadata that represets each download made
- db: where the processed files are stored as datasets (parquet files)

The folder structure looks like this:

```text
rb3.cachedir
├── raw
├── meta
└── db
```

The raw files are initially downloaded and stored in the `raw` folder.
These files are then processed and saved as parquet files in the `db` folder,
forming structured datasets that can be queried using the `rb3` package functions.
The data processing occurs in two stages: first, the raw files are transformed and
stored in the `input` layer within the `db` folder.
Next, the data undergoes further refinement and is saved in the `staging` layer,
also within the `db` folder.
The dataset cam be accessed using the function `rb3::template_dataset()`.

```{r template-dataset}
# Get the dataset for the template "b3-reference-rates"
template_dataset("b3-reference-rates")
#> FileSystemDataset with 47 Parquet files
#> 5 columns
#> refdate: date32[day]
#> curve_name: string
#> cur_days: int64
#> col1: double
#> col2: double
```

This function defaults to the `input` layer, but you can specify the `layer` argument
to access the `staging` layer if needed.

```{r template-dataset-input}
# Get the dataset for the template "b3-reference-rates" in the input layer
template_dataset("b3-reference-rates", layer = "staging")
#> FileSystemDataset with 47 Parquet files
#> 7 columns
#> curve_name: string
#> refdate: date32[day]
#> forward_date: date32[day]
#> cur_days: int64
#> biz_days: int64
#> col1: double
#> col2: double
```

We can observe that the dataset in the `input` layer has 5 columns, while the dataset
in the `staging` layer has 7 columns.
The datasets in the `staging` layer are enriched with formatted columns and additional
data.

## Accessing the data

In the previous sections we have seen how to download and process data using the
`fetch_marketdata()` function and how to access the downloaded data using the `template_dataset()`
function.
Each template has custom functions to access the data.
These functions have the suffix `_get()`.

- `cotahist_get()`: Retrieves historical stock market data.
- `futures_get()`: Retrieves futures settlement prices.
- `yc_brl_get()`: Retrieves the Brazilian nominal yield curve (PRE).
- `yc_ipca_get()`: Retrieves the Brazilian real interest rate curve (DIC).
- `yc_usd_get()`: Retrieves the FX-linked yield curve (DOC).

and many others.
For example, to access the data downloaded using the `b3-reference-rates` template,
you can use the `yc_brl_get()` function:

```{r yc-brazil-get}
# Get the Brazilian nominal yield curve (PRE)
yc_brl_get() |>
  filter(refdate == "2024-01-31") |>
  collect()
#> # A tibble: 257 × 7
#>    curve_name refdate    forward_date cur_days biz_days r_252  r_360
#>    <chr>      <date>     <date>          <int>    <int> <dbl>  <dbl>
#>  1 PRE        2024-01-31 2024-02-01          1        1 0.116 0
#>  2 PRE        2024-01-31 2024-02-07          7        5 0.112 0.115
#>  3 PRE        2024-01-31 2024-02-14         14        8 0.112 0.0906
#>  4 PRE        2024-01-31 2024-02-15         15        9 0.112 0.0953
#>  5 PRE        2024-01-31 2024-02-16         16       10 0.112 0.0994
#>  6 PRE        2024-01-31 2024-02-21         21       13 0.112 0.0983
#>  7 PRE        2024-01-31 2024-02-28         28       18 0.112 0.102
#>  8 PRE        2024-01-31 2024-02-29         29       19 0.112 0.104
#>  9 PRE        2024-01-31 2024-03-01         30       20 0.112 0.106
#> 10 PRE        2024-01-31 2024-03-04         33       21 0.112 0.101
#> # ℹ 247 more rows
```

The columns `r_252` and `r_360` have been renamed in the function `yc_brl_get()`.
This happens because the dataset `b3-reference-rates` attends to the three curves
PRE, DIC, and DOC, but the columns `col1` and `col2` have different meanings for each curve.
For this reason we strongly recommend using the custom functions to access the data instead of
using the `template_dataset()` function directly.

## Conclusion

The `rb3` package provides a comprehensive and efficient framework for accessing, processing,
and analyzing market data from B3 (the Brazilian stock exchange).
In this vignette, we explored the key functionalities of the package, including:

1.  **Downloading Market Data**: Using templates and the `fetch_marketdata()` function,
    we demonstrated how to download and process various types of market data, such as
    reference rates and futures settlement prices.
2.  **Data Storage and Organization**: We reviewed the structure of the `rb3.cachedir` folder,
    which organizes raw files, metadata, and processed datasets for efficient access and management.
3.  **Accessing Processed Data**: We showcased how to query the processed datasets using
    template-specific functions like `template_dataset()` and custom access functions such as
    `yc_brl_get()` and `futures_get()`.

By combining the power of templates, efficient data storage, and specialized query functions,
the `rb3` package simplifies the process of working with B3 market data.
Whether you are analyzing yield curves, futures prices, or other financial datasets, `rb3`
provides the tools needed to streamline your workflow and focus on generating insights.

We encourage you to explore the package further and adapt its functionalities to your specific
use cases in financial analysis.

