---
title: "Getting Started with rb3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rb3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `rb3` package provides tools for downloading, processing, and analyzing market data from B3 (the Brazilian stock exchange). This vignette will guide you through the basics of using the package to download various types of market data and perform common analyses.

```{r setup}
library(rb3)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(bizdays)
```

## Downloading Market Data

The main function for fetching market data is `fetch_marketdata()`. This function downloads data based on a template and parameter combinations, then processes the data into a structured database format.

### Historical Stock Market Data

You can download historical stock market data for specific years or date ranges:

```{r}
# Download yearly historical data
fetch_marketdata("b3-cotahist-yearly", year = 2020:2024)

# Download daily historical data for a specific date range
fetch_marketdata("b3-cotahist-daily", 
                 refdate = bizseq("2024-01-01", "2024-03-31", "Brazil/B3"))
```

### Futures Settlement Prices

To download settlement prices for futures contracts:

```{r}
# Download futures settlement prices for a date range
fetch_marketdata("b3-futures-settlement-prices", 
                 refdate = bizseq("2023-01-01", "2024-03-31", "Brazil/B3"))
```

### Reference Rates

You can download reference interest rates for various curves:

```{r}
# Download reference rates for multiple curves and dates
fetch_marketdata("b3-reference-rates",
                 refdate = bizseq("2023-01-01", "2024-03-31", "Brazil/B3"),
                 curve_name = c("DIC", "DOC", "PRE"))
```

### BVBG-086 Files

BVBG-086 files contain information about trading instruments:

```{r}
# Download BVBG-086 files for a date range
fetch_marketdata("b3-bvbg-086",
                 refdate = bizseq("2023-01-01", "2024-03-31", "Brazil/B3"))
```

## Querying Historical Stock Data

After downloading data, you can query it using specialized functions. The `cotahist_get()` function retrieves historical stock market data.

### Analyzing Equities

```{r}
# Get equity data
equities <- cotahist_get() |>
  cotahist_filter_equity() |>
  collect()

# Calculate daily returns for a specific stock
stock_data <- equities |>
  filter(symbol == "PETR4") |>
  arrange(refdate) |>
  mutate(daily_return = (price_close / lag(price_close)) - 1)

# Plot closing prices
ggplot(stock_data, aes(x = refdate, y = price_close)) +
  geom_line() +
  labs(title = "PETR4 Closing Prices",
       x = "Date",
       y = "Price (BRL)")

# Check for missing data points
date_range <- equities |>
  summarise(min = min(refdate), max = max(refdate))

check_dates <- tibble(refdate = bizseq(date_range$min, date_range$max, "Brazil/B3"))

missing_data <- equities |>
  filter(symbol == "PETR4") |>
  full_join(check_dates) |>
  arrange(refdate) |>
  filter(is.na(price_close))
```

### Analyzing BDRs (Brazilian Depositary Receipts)

```{r}
# Get BDR data
bdrs <- cotahist_get() |>
  cotahist_filter_bdr() |>
  collect()

# Find most frequently traded BDRs
top_bdrs <- bdrs |>
  group_by(symbol) |>
  count() |>
  arrange(desc(n)) |>
  head(10)

# Analyze a specific BDR
bdr_data <- bdrs |>
  filter(symbol == "AAPL34") |>
  arrange(refdate)

# Plot trading volume
ggplot(bdr_data, aes(x = refdate, y = trade_quantity)) +
  geom_line() +
  labs(title = "AAPL34 Trading Volume",
       x = "Date",
       y = "Number of Trades")
```

## Analyzing Futures Data

The `futures_get()` function allows you to access futures settlement prices.

```{r}
# Get futures data for interest rate contracts
di1_futures <- futures_get() |>
  filter(commodity == "DI1") |>
  collect() |>
  mutate(
    maturity_date = maturity2date(maturity_code),
    fixing = following(maturity_date, "Brazil/ANBIMA"),
    business_days = bizdays(refdate, maturity_date, "Brazil/ANBIMA"),
    adjusted_tax = (100000 / price)^(252 / business_days) - 1
  ) |>
  filter(business_days > 0)

# Plot interest rates for specific contracts
di1_futures |>
  filter(symbol %in% c("DI1F25", "DI1F26")) |>
  ggplot(aes(x = refdate, y = adjusted_tax, color = symbol, group = symbol)) +
  geom_line() +
  geom_point() +
  labs(
    title = "DI1 Future Rates - Nominal Interest Rates",
    x = "Date",
    y = "Interest Rates",
    color = "Symbol"
  ) +
  scale_y_continuous(labels = scales::percent)
```

## Working with Yield Curves

The `rb3` package provides specialized functions for working with yield curves:

- `yc_brl_get()`: Retrieves the Brazilian nominal yield curve (PRE)
- `yc_ipca_get()`: Retrieves the Brazilian real interest rate curve (DIC)
- `yc_usd_get()`: Retrieves the FX-linked yield curve (DOC)

### Analyzing the PRE Curve (Nominal Interest Rates)

```{r}
# Get PRE curve data
pre_curve <- yc_brl_get() |>
  filter(forward_date < "2035-01-01") |>
  collect()

# Plot yield curves for different dates
ggplot(
  pre_curve,
  aes(
    x = forward_date,
    y = r_252,
    group = refdate,
    color = factor(refdate)
  )
) +
  geom_line(linewidth = 1) +
  labs(
    title = "Nominal Yield Curves for Brazil",
    subtitle = "Built using interest rates future contracts",
    x = "Forward Date",
    y = "Annual Interest Rate",
    color = "Reference Date"
  ) +
  scale_y_continuous(labels = scales::percent)
```

### Analyzing the DIC Curve (Real Interest Rates)

```{r}
# Get DIC curve data
ipca_curve <- yc_ipca_get() |>
  collect()

# Plot real interest rate curves
ggplot(
  ipca_curve |> filter(biz_days > 21, biz_days < 1008),
  aes(
    x = forward_date,
    y = r_252,
    group = refdate,
    color = factor(refdate)
  )
) +
  geom_line(linewidth = 1) +
  labs(
    title = "Real Interest Rate Curves for Brazil",
    subtitle = "Built using interest rates future contracts",
    x = "Forward Date",
    y = "Annual Interest Rate",
    color = "Reference Date"
  ) +
  scale_y_continuous(labels = scales::percent)
```

### Calculating Break-even Inflation

By comparing nominal and real yield curves, you can derive the implied inflation expectation:

```{r}
# Load PRE and DIC curves
pre <- yc_brl_get() |>
  select(refdate, forward_date, r_nominal = r_252) |>
  collect()

ipca <- yc_ipca_get() |>
  select(refdate, forward_date, r_real = r_252) |>
  collect()

# Calculate break-even inflation
break_even <- inner_join(pre, ipca, by = c("refdate", "forward_date")) |>
  mutate(break_even = r_nominal - r_real)

# Plot break-even inflation
ggplot(
  break_even,
  aes(
    x = forward_date,
    y = break_even,
    group = refdate,
    color = factor(refdate)
  )
) +
  geom_line(linewidth = 1) +
  labs(
    title = "Implied Inflation (Break-even) from Yield Curves",
    subtitle = "Calculated as PRE (nominal) minus DIC (real)",
    x = "Forward Date",
    y = "Annual Implied Inflation Rate",
    color = "Reference Date"
  ) +
  scale_y_continuous(labels = scales::percent)
```

## Advanced Example: Forward Rates

Calculate 10-year forward rates from DI1 futures contracts:

```{r}
# Get futures data
futures_data <- futures_get() |>
  filter(symbol %in% c("DI1F24", "DI1F34")) |>
  mutate(
    maturity_date = maturity2date(maturity_code) |>
      following("Brazil/ANBIMA"),
    business_days = bizdays(refdate, maturity_date, "Brazil/ANBIMA")
  ) |>
  collect()

# Calculate forward rates
business_days_diff <- futures_data |>
  select(refdate, symbol, business_days) |>
  pivot_wider(names_from = symbol, values_from = business_days) |>
  mutate(du = DI1F34 - DI1F24) |>
  select(refdate, du)

forward_rates <- futures_data |>
  select(refdate, symbol, price) |>
  pivot_wider(names_from = symbol, values_from = price) |>
  inner_join(business_days_diff, by = "refdate") |>
  mutate(fwd = (DI1F24 / DI1F34)^(252 / du) - 1) |>
  select(refdate, fwd) |>
  na.omit()

# Plot forward rates
ggplot(forward_rates, aes(x = refdate, y = fwd)) +
  geom_line() +
  labs(
    x = "Date", 
    y = "Forward Rates",
    title = "Historical 10Y Forward Rates",
    caption = "Source B3 - package rb3"
  ) +
  scale_y_continuous(labels = scales::percent)
```

## Conclusion

The `rb3` package provides a comprehensive set of tools for working with Brazilian market data. This vignette has shown you how to:

1. Download various types of market data using `fetch_marketdata()`
2. Query and analyze historical stock data
3. Work with futures contracts and calculate implied rates
4. Analyze yield curves and derive market-implied metrics like break-even inflation
5. Calculate forward rates from futures prices

For more detailed examples and analyses, check out the other vignettes:

- "Fetching Market Data with rb3"
- "Fetching Historical Yield Curves from B3"
- "How to Compute Historical Rates from B3 Future Prices"

Remember that all downloaded data is stored in a local database managed by the `rb3` package, making it readily available for queries without needing to download it again.